---
title: "Elementary considerations on an R/HSDS interface"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Elementary considerations on an R/HSDS interface}}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

```{r setup,include=TRUE,results="hide",message=FALSE,warning=FALSE}
library(rhsdsclient)
library(SummarizedExperiment)
```

# Introduction

The Bioconductor project makes significant use of HDF5 as an out-of-memory
storage format for large numerical arrays generated in bioinformatic applications.

HSDS is the [Highly Scalable Data Service](https://www.hdfgroup.org/solutions/highly-scalable-data-service-hsds/) produced by the HDF Group, that
"brings HDF5 to the cloud".

This document addresses work on interfacing R and HSDS.  The specific use
case of interest is interrogating matrices stored in HSDS to retrieve
submatrices, using natural idioms in R including processes that
work with Bioconductor SummarizedExperiment instances.

# Using the HSDS CLI

To get some context, if we have installed the python
package [h5pyd}(https://github.com/HDFGroup/h5pyd) appropriately (currently
I am using version 0.8.3), we could use the `hsls` command to
learn about a test dataset.
```
> system("hsls -r -e http://hsdshdflab.hdfgroup.org /shared/tall.h5", intern=TRUE)
 [1] "/ Group"                                                   
 [2] "/g2 Group"                                                 
 [3] "/g2/dset2.2 Dataset {3, 5}"                                
 [4] "/g2/dset2.1 Dataset {10}"                                  
 [5] "/g1 Group"                                                 
 [6] "/g1/g1.2 Group"                                            
 [7] "/g1/g1.2/g1.2.1 Group"                                     
 [8] "/g1/g1.2/g1.2.1/slink    SoftLink {somevalue}"             
 [9] "/g1/g1.2/extlink         ExternalLink {somepath//somefile}"
[10] "/g1/g1.1 Group"                                            
[11] "/g1/g1.1/dset1.1.2 Dataset {20}"                           
[12] "/g1/g1.1/dset1.1.1 Dataset {10, 10}"                       
```

This tells us that the HSDS deployment at `http://hsdshdflab.hdfgroup.org`
has an endpont `/shared/tall.h5/g2/dset2.2` that is an array with shape `(3,5)`.

# A small example with rhsdsclient

## Some R code

We'll continue working with the deployment at `http://hsdshdflab.hdfgroup.org`.
Methods HSDSSource, HSDSFile, HSDSDataset are used to set up interrogations
of HDF content.

```{r lktal}
src = HSDSSource("http://hsdshdflab.hdfgroup.org", "hsds")
tall = HSDSFile(src, "/shared/tall.h5")
tall
listDatasets(tall)
HSDSDataset(tall, "/g1/g1.1/dset1.1.1")
```

Bracket methods have been defined to extract
numerical content.
```{r lkbra}
HSDSDataset(tall, "/g1/g1.1/dset1.1.1")[1:4,1:4]
```

## The GET request

The last R call produced the following request (broken up
across multiple lines for readability):

```
GET http://hsdshdflab.hdfgroup.org/datasets/
   d-d38053ea-3418fe27-22d9-478e7b-913279/
   value?domain=/shared/tall.h5&select=[0:4:1,0:4:1]
```

The task of the rhsdsclient package is to take
matrix-oriented requests, which may involve rownames
and colnames, and translate them into valid requests,
potentially reorganizing results so that the
expectations of the R user are met.
